### 2023-01-12 Manuscript/Draft1.Rmd
Paper and VAST code updates

--Stratified mean vs VAST with and without covariates
[[could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators]] 
[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]
**BL: Chris, please check paragraph below for intro sentence and help fill in X, Y and Z**

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. There are a number of commonly used model-based estimates such as X, Y and Z. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance]. VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

>>>>

--Stratified mean vs VAST with and without covariates
[[could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators]] 
[[may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations]]

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. Model-based abundance estimates are often derived from model framewroks such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model [@thorson2019guidance]. VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

### 2023-01-12 Manuscript/Draft1.tex
fairly minor paper updates
Paper and VAST code updates

{\centering \includegraphics[width=0.95\linewidth]{Images/Strata}

@@

--Stratified mean vs VAST with and without covariates
{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]}
{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}
\textbf{BL: Chris, please check paragraph below for intro sentence and help fill in X, Y and Z}

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. There are a number of commonly used model-based estimates such as X, Y and Z. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019). VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.

>>>>

{\centering \includegraphics[width=0.95\linewidth]{Images/Strata} \includegraphics[width=0.95\linewidth]{Images/Excluded_strata}

@@

--Stratified mean vs VAST with and without covariates
{[}{[}could use an intro sentence here to distinguish design-based (stratified mean) and model-based (VAST) estimators{]}{]}
{[}{[}may want to highlight the potential of VAST with covariates to address the climate change issue here as the reason we are doing these simulations{]}{]}

Stock assessment scientists use samples from the bottom trawl survey to estimate the abundance for each stock stock. There are a number of approaches to obtaining abundance estimates ranging from traditional design-based estimates such as the stratified mean to model-based estimates that range in complexity. The stratified mean calculates the geometric mean catch per tow and does not account for spatial variation in samples or allow the inclusion of environmental covariates. Model-based abundance estimates are often derived from model framewroks such as General Linear Models (GLM), General Additive Model (GAM), and General Linear Mixed Models (GLMM) that can incorporate both random and mixed effects. The Vector Autoregressive Spatio-Temporal (VAST) model is a spatial delta-generalized linear mixed model (Thorson 2019). VAST is a spatio-temporal statistical framework that models both abundance (biomass) and probability of occurrence (presence/absence). If desired, VAST also allows users to include covariate data to better inform the model. Covariates can be static (eg. habitat preferences) or dynamics (eg. temperature). We explore whether including environmental predictors can help inform models and provide better abundance estimates, which is particularly relevant as climate change progresses. The stratified mean calculations are straightforward and quick, while VAST models require numerous user inputs and take on the order of hours to complete. We compare the yearly estimated of abundance obtained from the stratified mean to estimates obtained from the Vector-Autoregressive Spatio-temporal (VAST) model.
